Hippocampal segmentation with MAGeT

### 2013-01-22 paper.Rnw
Updated introduction to include a clearer argument for MAGeT brain as addressing segmentation with small atlas sets

Compounding each of these problems is the significant neuroanatomical
variability in the hippocampus throughout the course of aging, development, and
neuropsychiatric disorders \citep{Mouiha2011b}.  Additionally, it may be
necessary to use several different hippocampal definitions or, in fact, make
specific modifications in the course of research. For example, Poppenk et al.
\citep{Poppenk2011} found that subdividing the hippocampus into anterior and
posterior regions resulted in a predictive relationship between volume
difference of those regions and recollection memory performance.  Thus, while
manual segmentation of the hippocampus is a necessary technique, to researchers
or clinicians who do not have access to the needed human expertise its use may
be infeasible.

%  Automated methods 
%
%  Heckemann2006 - 
%  Heckemann2011 -  Automatic morphometry in AD and MCI 
%    - (MAPER). multi-atlas. 30 atlases, entire ADNI screening  
%  Collins2010 - .. seg of HC and AM by augmenting ANIMAL
%    - multi-atlas 80 atlases, atlas selection (NMI)
%  Aljabar2009 - atlas selection and effect on accuracy
%    - 
%  Leung2010 - X-sectional and longitudinal HC vol in MCI, AD
%    - (MAPS), multi-atlas + STAPLE. 55 atlases. baseline+1yr of ADNI seg.
%    - atrophy rates
%  Lotjonen2010 - Fast and robust
%    - IBSR/ADNI.  
%
%
%  Hippocampal Volume
%
%  Mouiha2011b - Hippocampal atrophy rates in AD
%    - compare SNT vs FreeSurfer on ADNI
%
Automated segmentation techniques overcome the need for human expertise by
performing segmentations computationally.  A popular class of automated
methods, {\it multi-atlas-based segmentation}, rely on a set of expertly
labeled neuroanatomical atlases. Each atlas is warped to fit a subject's
neuroanatomy using nonlinear registration
techniques \citep{Collins1995,Klein2009}.  Atlas labels are then transformed by
this warping and a {\it label fusion} technique, such as voxel-wise voting, is
used to merge the competing label definitions from each atlas into a final
segmentation for a subject.

Many descriptions of multi-atlas-based segmentation algorithms report relying
on an atlas library containing anywhere between 30 and 80 expertly labeled
brains \citep{Heckemann2011,Collins2010,Aljabar2009,Leung2010,Lotjonen2010}.  As
noted, the production of an atlas library requires significant manual effort,
and is limited since the choice of atlases or segmentation protocol may not
reflect the underlying neuroanatomical variability of the population under
study or be suited to answer the research questions at hand.

In this paper we propose an automated segmentation method to address the above
issues of existing multi-atlas-based methods. Principally, our method aims to
dramatically reduce the number of manually labelled atlases needed (under 10).
This is achieved by using the small atlas library to bootstrap a much larger
"template library", which is then used to segment the subjects in a similar
fashion to basic multi-atlas segmentation. This approach has the additional
advantage of using the unique subject population on hand to initialize the
segmentation process and improve accuracy.

The essential insight of generating a template library is not new.  Heckemann
\citep{Heckemann2006} compared generating a template library from a single atlas
to standard multi-atlas segmentation and found poor performance and so deemed
the approach as inviable.  The LEAP algorithm \citep{Wolz2010} proceeds by
iteratively segmenting the unlabelled image most similar to the atlas library
images and then incorporating the now-labelled image into the atlas library,
but requires 30 starting atlases.  The novelty of our method is to demonstrate
the possibility of producing comparable segmentation accuracy to these and
other multi-atlas-based methods while using significantly fewer manually
created atlases.

In our previous work \citep{Chakravarty2012}, we applied MAGeT brain to
segmentation of the human striatum, globus pallidus, and thalamus using a
single histologically-derived atlas. The main contribution of this paper is to
extend our approach to the human hippocampus and perform a thorough validation
over a range of atlas and template library sizes, which was not done in our
previous work.  Due to the small number of atlases required, our method can
easily accommodate different hippocampal definitions. Our aim is not to improve
on segmentation accuracy beyond existing methods, but instead to provide a
method that trades off manual segmentation expertise for computational
processing time while providing sufficient accuracy for clinical and research
applications.

>>>>

Compounding each of these problems is the significant neuroanatomical
variability in the hippocampus throughout the course of aging, development, and
neuropsychiatric disorders \citep{Mouiha2011b}.  The result is that existing
hippocampal atlases available to a researcher may not accurately represent
neuroanatomy of a specific population under study.  Additionally, in the course
of a research or clinical study, it may be necessary to make adjustments to
hippocampal definition as a means of hypothesis testing.  For example, Poppenk
\citep{Poppenk2011} found that subdividing the hippocampus into anterior and
posterior regions resulted in a predictive relationship between volume
difference of those regions and recollection memory performance.  Making
modifications to an set of atlas library segmentations may require the same
level of expertise and manual effort needed to create the segmentations from
scratch.

%  Automated methods 
%
%  Heckemann2006 - 
%  Heckemann2011 -  Automatic morphometry in AD and MCI 
%    - (MAPER). multi-atlas. 30 atlases, entire ADNI screening  
%  Collins2010 - .. seg of HC and AM by augmenting ANIMAL
%    - multi-atlas 80 atlases, atlas selection (NMI)
%  Aljabar2009 - atlas selection and effect on accuracy
%    - 
%  Leung2010 - X-sectional and longitudinal HC vol in MCI, AD
%    - (MAPS), multi-atlas + STAPLE. 55 atlases. baseline+1yr of ADNI seg.
%    - atrophy rates
%  Lotjonen2010 - Fast and robust
%    - IBSR/ADNI.  
%  Mouiha2011b - Hippocampal atrophy rates in AD
%    - compare SNT vs FreeSurfer on ADNI
%
Automated segmentation techniques overcome the need for human expertise by
performing segmentations computationally.  A popular class of automated
methods, {\it multi-atlas-based segmentation}, use a series of expertly
segmented MRIs (an atlas library) to act as templates against which a target
image is labelled
\cite{Aljabar2009,Collins2010,Heckemann2011,Leung2010,Lotjonen2010,Leung2010,Wolz2010}.
In the basic method, each atlas is warped to fit a subject's neuroanatomy using
nonlinear registration techniques \citep{Collins1995,Klein2009}.  Atlas labels
are then transformed by this warping and a {\it label fusion} technique, such
as voxel-wise voting, is used to merge the competing label definitions from
each atlas into a final segmentation for a subject.

Many descriptions of multi-atlas-based hippocampal segmentation algorithms
report relying on an atlas library containing anywhere between 30 and 80
expertly labeled images
\citep{Heckemann2011,Collins2010,Aljabar2009,Leung2010,Lotjonen2010} in order
to achieve performance approach that of a manual rater.  As already noted, the
production of an atlas library requires significant manual effort and
expertise, as does the modification of an existing atlas library in order to
better suit a target population or investigate a specific research hypothesis.
Therefore, in this paper we tackle the problem of producing accurate
segmentations using small numbers of manually segmented atlases.

In this paper we propose an extension to the basic multi-atlas-based
segmentation scheme to address this issue.  Principally, we explore the
possibility of using a small atlas library to bootstrap a much larger
\emph{template library} composed of images taken from the target population.
The template library is then used to segment the subjects in a similar fashion
to basic multi-atlas segmentation: by label propagation and label fusion.  This
intuition driving this approach is to leverage the unique neuroanatomy of
subject population on hand to initialize the segmentation process and improve
accuracy over direct propagation from the atlas library to unlabelled subjects.

The essential insight of generating a template library is not new.  Heckemann
\citep{Heckemann2006} briefly explored "indirect" segmentation by taking a
single atlas and propagating those labels to intermediate subjects before
fusing them in a target image space.  Heckemann compared indirect single-atlas
segmentation to direct multi-atlas segmentation found the indirect approach
performed worse.  In this paper we continue the same line of investigation but
explore the performance when using multiple atlases, as well as several
different registration and fusion methods.

The LEAP algorithm \citep{Wolz2010} is another example of indirect segmentation
previously explored.  LEAP proceeds by iteratively segmenting unlabelled images
most similar to the atlas library images and then incorporating the
now-labelled images into the atlas library for future iterations.  30 manually
labelled atlases are required to start the process.  The novelty explored in
our current work is demonstrate the viability of acheiving comparable
segmentation accuracy while using significantly fewer manually created atlases.

In previous work \citep{Chakravarty2012}, we applied MAGeT brain to
segmentation of the human striatum, globus pallidus, and thalamus using a
single histologically-derived atlas. The main contribution of this paper is to
extend our approach to the human hippocampus - a much more difficult structure
to segment -- and perform a thorough validation over a range of atlas and
template library sizes, registration methods, and fusion methods.  Our aim in
this work is not necessarily to improve on segmentation accuracy beyond
existing methods, but instead to provide a method that trades off manual
segmentation expertise and time for computational processing time while still
providing sufficient accuracy for clinical and research applications.
